# The Definite Integral

## Introduction

Here is some code that approximates the area under $f(x)=x^2$ on the interval $[a,b]$ using **right hand endpoints** on 100 subintervals. (How would you change this code to use **left hand endpoints** instead?)

```{r}
f=makeFun(x^2 ~ x)
a = 0
b = 3
num = 100
### the approximation of f(x) on interval [a,b]
base = (b-a)/num
points = seq(from=a+base, to=b, by=base)
heights = f(points)
areas = base*heights
sum(areas)
```

## Activities

### Drive My Car

A car is driving at a rate of $v(t)=-t+5$. Here is a plot of the velocity.

```{r  echo=FALSE, fig.keep='last'}
plotFun(-t+5~t, xlim=range(0,12),xlab='t',ylab='v(t)')
plotFun(0*t~t, add=TRUE)
```


1. Use RStudio to approximate these definite integrals using 1000 subintervals.
  a. $\displaystyle{\int_0^5 v(x)dx}$
  b. $\displaystyle{\int_0^8 v(x)dx}$
  c. $\displaystyle{\int_0^{12} v(x)dx}$

2. When does the car **turn around**? 

3. When does the car **return to where it started**? 


## Accelerate My Car

A car is traveling at 55 feet per second at time $t=0$. The plot below shows the car's **acceleration** on the interval $[0,18]$.

```{r echo=FALSE, fig.align="center", fig.keep="last"}
a=makeFun(ifelse(x<2,8,ifelse(x<10,-2*x+12,ifelse(x<14,-8,ifelse(x<18,4*x-64,8))))~x)
plotFun(a(t)~t,xlim=range(0,18))
plotFun(0*t~t, add=TRUE)
```

1. Is the velocity increasing or decreasing at $t=2$? $t=15$? 
2. What happens when $t=6$?
3. When else (besides $t=0$) is the velocity 55 feet per second?
4. What is the **maximum** velocity, and when does it occur? 
5. What is the **minimum** velocity, and when does it occur?


## Making Cupcakes

The marginal cost of making cupcakes is given by $$K'(q)=\frac{1}{100}q^2+\frac{1}{5}q+1,$$ 
where $q$ is in dozens of cupcakes. 

1. Set up a definite integral to calculate $K(30)-K(0)$, which is the change in costs when we create 30 dozen cupcakes.
2. Estimate this definite integral using *RStudio*. 
3. If our fixed costs are \$1500, what is the **total cost** to make 30 dozen cupcakes? 

## Area Between Curves

Below is a graph of both $\sin(x)$ and $\cos(x)$ on the interval $[0,2\pi]$. 



```{r fig.align="center", fig.keep='last'}
plotFun(sin(x)~x,xlim=range(0,2*pi),ylim=range(-1.1,1.1),xlab="",ylab="")
plotFun(cos(x)~x,add=TRUE)
```

1. Write a **sum of integrals** which represents the \emph{area} enclosed by the two curves on the interval $[0,2\pi]$.   
Hint: $\sin(x) = \cos(x)$ at $x=\pi/4$ and $x=5\pi/4$.
2. Use *RStudio* to estimate the area between these two curves.




