# RStudio Quick Referrence

```{r, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(mosaic))
```
    
## Common Functions

| function | R command | comments |
|-----|-----|-----|
| $x^2$ | `x^2` | |
| $4x^2 -7x + 3$ | `4*x^2 - 7*x + 3` | You must **always** include the `*` when you multiply |
| $\sqrt{x}$| `sqrt(x)` or `x^(1/2)` ||
| $\sqrt[3]{x}$ | `x^(1/3)` ||
| $\sin(x)$ | `sin(x)` ||
| $\cos(x)$ | `cos(x)` ||
| $e^x$ | `exp(x)` | `exp` is a **function** (like `sin` and `cos`), so you do NOT need a `^` |
| $\mbox{ln}(x)$ | `log(x)` | RStudio uses the natural logarithm as the default |
| $\log(x) = \log_{10}(x)$ | `log10(x)` | Humans default to "log base 10." RStudio does not!
| $\log_{b}(x)$ | `log(x,b)` | You can use any number $b$ as the base for your logarithm


## Defining your own function

The `makeFun` command is part of the mosaic package.




| desired function | R command | comments |
|-----|-----|-----|
| $f(x) = x^2 +5x + 6$ | `f = makeFun(x^2 + 5*x + 6 ~ x)` | remember your `*` signs! |
| $g(x) = \sin^2(x) - \frac{1}{2}$ | `g = makeFun(sin(x)^2 - 1/2 ~ x)` | note the placement of `^2` |
| $P(t) = 5 e^{.25 t}$ | `P = makeFun(5 * exp(0.25 * t) ~ t)` | `exp` does not need a `^` |
| $Q(t) = 12.38 (1.041)^t$ | `Q = makeFun(12.38 * (1.041)^t ~ t)` ||


Notes:

* The general syntax is `makeFun( OUTPUT ~ INPUT )` where the OUTPUT is a function of the INPUT.
* You must **assign this to a variable** so that you can use it later. So your command must **start with** `f = ...` so that you get a function named `f`.
* The variable (`x` or `t`, etc) that appears in the OUTPUT **must match** the INPUT variable that appears after the tilde `~`. The the command `makeFun(x^2 ~ t)` won't work because the `x^2` is not a function of `t`.


## Creating Lists of Data

Use `c()` to "combine" some values into a list. Assign the list to a variable so that you can use it later.

```{r}
my_primes = c(2, 3, 5, 7, 11, 13, 17, 19, 23, 29)
my_primes

my_data = c(5, -2, 7, 3, -10, 15)
my_data


```

Use `seq()` to generate a sequential list of values in a specified range. The optional third parameter tells RStudio to increment by that value instead of incrementing by 1.

```{r}

my_seq1 = seq(1,10)
my_seq1

my_seq2 = seq(5, 6, 0.25)
my_seq2

my_seq3 = seq(5, 6, 0.1)
my_seq3

```



## Plotting Data and Plotting Functions


Use `plotPoints` to plot a list of $y$-axis data versus a list of $x$-axis data. Separate these lists with a tilde `~`. 

```{r}

X = seq(1:5)
Y = X^2
plotPoints(Y ~ X)

```


```{r}

x_data = c(2,3,5,8,13)
y_data = c(2,4,8,4,2)
plotPoints(y_data ~ x_data)

```

Use `plotFun` to plot a function. 

* As with `plotPoints`, we use a tilde `~` to separate the output variable (the function) from the input variable.
* The optional `xlim` argument specifies the domain for the plot.

```{r}
plotFun(sin(x) ~ x)

plotFun(sin(x) ~ x, xlim=c(0, 2*pi))
```

We can layer one plot on top of another by using the `add=TRUE` argument.

* You only need to specify the `xlim` for the **first plot** because the second one is added to the first.
* It's helpful to change the color for subsequent plots. Use the argument `col="red"` using any color name you like (RStudio knows a lot of colors!)


```{r}
plotFun(2 * sin(x) ~ x, xlim=c(0, 2*pi))
plotFun(cos(3*x) ~ x, add=TRUE, col='green')
```

We can label the axes using `xlab` and `ylab`. When you are layering two plots, you must specify these labels in the first function that you plot!


```{r}
plotFun(2 * sin(x) ~ x, xlim=c(0, 2*pi), xlab='time', ylab='displacement')
plotFun(cos(3*x) ~ x, add=TRUE, col='green')
```

We can also combine plots of data and functions using `add=TRUE`.


```{r}

time_data = seq(0,50, 10)
pop_data = c(6.82, 9.63, 14.3, 20.48, 28.94, 42.25)

plotPoints(pop_data ~ time_data, xlab='years since 1950', ylab='population (millions)')

P = makeFun( 6.82 * (1.037)^t ~ t)

plotFun(P(t) ~ t, add=TRUE, col='orange')


```

## Fitting a Function to Data

The `fitModel` command is part of the mosaic package. Here is how you would fit a linear model to data

```{r}

xdata = c(1,2,3,4,5)
ydata = c(5.3, 7.7, 10.1, 11.3, 12.0    )

fitModel(ydata  ~ A* xdata  + B)

```
This output says that the best fitting linear function has slope $A=1.70$ and $y$-intercept $B=4.18$. So let's create the corresponding linear function and plot it with the data.

```{r}
f = makeFun(1.70 *  x + 4.18 ~ x)

plotPoints(ydata ~ xdata)
plotFun(f(x) ~ x,  add=TRUE, col='orange')
```

### Fitting a Power Function to Data

Here are the steps to fit `xdata` and `ydata` to a power function $f(x) = c x^p$. 

1. Take the log of both `xdata` and `ydata`
2. Fit a linear function $y = A x + B$ to the resulting data
3. The original constants are `p = A` and `c = exp(B)`

Here is an example

```{r}
f = makeFun(4.3 * x^(1.7)~x)

xdata = c(1, 3, 5, 7, 9)
ydata = c(7, 33, 65, 110, 180)

logxdata = log(xdata)
logydata = log(ydata)

fitModel(logydata ~ A * logxdata + B)

```

This tells us that $A=1.45$ and $B=1.92$. We solve for $c$ and plot the function with the data

```{r}
p = 1.45
p
c = exp(1.92)
c

plotPoints(ydata ~ xdata)
plotFun(6.82 * x^1.45 ~ x, add=TRUE, col='orange')

```

### Fitting a Model by Specifying Starting Values

Sometimes `fitModel` fails to find a good model for the data. Here are two things that can happen.

* The command  `fitModel` fails,complaining about a "singular gradient." This means that `fitModel` picked a bad starting point, and can't figure out how to improve it's initial guess.
* The command `fitModel` does find a function, but when we plot it, we realize that it's not what we want. 

Here is an example of `fitModel` returning a function that we don't like. Let's load in the `utilities.csv' data set and look at the first few rows.

```{r}
utilitydata = fetch::fetchData("utilities.csv")
head(utilitydata)
```

Now let's plot temperature versus month

```{r}
plotPoints(temp ~ month, data=utilitydata)
```

We have ploted monthly temperature data for multiple years. Let's try to fit a periodic function.

```{r}
M = fitModel(temp ~ A * sin(w*(month + p))+C, data=utilitydata)
coef(M)

f = makeFun(-28.7 * sin(-6.8*(t + 56.2))+47.2 ~ t)
plotFun(f(t)~t, add=TRUE, col='red')
```

This is not what we wanted! It looks like `fitModel` was confused by the multiple data points for each month. So the frequency `w` is completely wrong! We can correct this by specifying an initial guess for `w`. We think that the period should be 12 months, so we want to start with `w=2*pi/12`.

```{r}
M = fitModel(temp ~ A * sin(w*(month + p))+C, data=utilitydata, start=list(w=2*pi/12))
coef(M)

f = makeFun(-28.7 * sin(0.5*(t -22.8))+47.2 ~ t)
plotFun(f(t)~t, add=TRUE, col='orange')
```





That looks much better!

## Plotting Surfaces in 3D

We plot functions $z = f(x,y)$ using the `plotFun()` command. We specify the independent variables with the syntax `~ x&y`. By default, RStudio will create a **contour plot** of this function and include **shading** to help visualize the values of the function.


```{r}
f = makeFun(x^2 - y^2 ~ x&y)

plotFun(f(x,y) ~ x&y)

```

You can add more contours using the `levels` parameter, and turn off the shading with the `filled` parameter.

```{r}
plotFun(f(x,y) ~ x&y, levels = seq(-1 ,1, .2), filled = FALSE)

```


You can change the plot range using `xlim` and `ylim`. 

```{r}
plotFun(f(x,y) ~ x&y, xlim=c(-20,20), ylim=c(-20,20))
```

You can create a 3D plot by setting `surface=TRUE`.

```{r}
plotFun(f(x,y) ~ x&y, xlim=c(-20,20), ylim=c(-20,20), surface=TRUE)
```


## Troubleshooting
Without fail, you will run into problems. Sometimes RStudio will give you helpful error messages, but sometimes they are obtuse. Here's a list of some common errors and what might be wrong:

- _Nothing ran, all I got was a plus sign..._ You probably forgot some parentheses. Try the following:
   - Enter a parenthesis on that line to see if this closes it out.
   - Press the Escape key to get to a new line. 


* _Unexpected symbol in..._ You may have forgotten a multiplication sign or otherwise gave RStudio a symbol it cannot parse. Check your line of code again to see if you have all your symbols.

* _Object not found..._ You may not have defined that variable (or function) yet. Check your environment tab to see if it has the object you're looking for (or if you've called it something different than you remember!). You also may be trying to use something from a data set that you have not included as a parameter. See if you need to add in the parameter `data=....`

* _Could not find function..._ RStudio doesn't know what function you're using. Check your capitalization and that you have the right packages installed.

* _... number of items to replace is not a multiple of replacement length ..._ Your input variable doesn't match the name variable you used in the output when you called `makeFun`. Maybe you used `x` on one side of the tilde `~` and `t` on the other side? 

* _Argument "name" is missing, with no default..._
You're probably trying to take a derivative or antiderivative without the mosaicCalc package. Go back to the packages section of the introduction for how to fix this.

* _I can't find my data!_  You may have fetched it but did not give it a name. Go back to your fetch command and make sure you have assigned it a variable.

