
# Gradient Search


## Gradient Search Example

Let's use gradient search to find the maximum of the function 
$$
f(x,y) = - x^4 -  x^3 + 10 x y + 2y - 8 y^2 
$$
whose partial derivatives are
$$
\frac{\partial f}{\partial x} = -4 x^3 -3 x^2  + 10 y
\quad \mbox{and} \quad
\frac{\partial f}{\partial x} = 10 x + 2 - 16 y
$$


First, you define the partial derivatives and then choose your starting point `(newx, newy)`. In this case, we start at `(1,1)`.

\medskip

```{r}
partialx = makeFun( -4*x^3 -3*x^2 + 10*y ~ x&y)
partialy = makeFun(10*x + 2 - 16*y ~ x&y)
newx = 1
newy = 1
```

\medskip

Next, you **repeatedly**  run the following code block, which updates the current point by moving `0.1` times the gradient vector. This takes a small step in the uphill direction.

```{r eval=FALSE}
newx = newx + 0.1*partialx(newx, newy)
newy = newy + 0.1*partialy(newx, newy)
# print new partial derivatives
c(partialx(newx, newy), partialy(newx, newy))
# print new point
c(newx, newy)
```


Repeatedly run this code block until the partial derivatives are essentially zero (at least two zeros after the decimal point). Congrats! You have found your local maximum.

You will end at the point `(1.04,0.77)`. But note that starting at another initial point might take you to a different local maximum!


## Activities

### Finding a Local Minimum

1. How should I change to the code if I want to find a **local minimum** instead of a local maximum? (Hint: we want to take a small step downhill.)
2. Use gradient search to find a **local minimum** of 
$$f(x,y) = x^2 + 2 x y + 3 x + 4 y + 5 y^2.$$
