# 2D Optimization

## Summary

The **critical points** of $f(x,y)$ are the points $(a,b)$ where both $f_x(a,b)=0$ and $f_y(a,b)=0$.


There are three ways to determine whether a critical point $(a,b)$ is a minimum, a maximum, or a saddle point.
  
1. The second derivative test relies on the value
$$
D = f_{xx} f_{yy} - (f_{xy})^2.
$$
   * If $D > 0$ and $f_{xx} > 0$ then $(a,b)$ is a local minimum.
   * If $D > 0$ and $f_{xx} < 0$ then $(a,b)$ is a local maximum
   * If $D < 0$ then $(a,b)$ is a saddle point
   * If $D=0$ then the test fails
   
2. Create a contour plot for $f(x,y)$ on a very small neighborhood of $(a,b)$. Use the contours to figure out whether it is a local max, a local min, or a saddle point.

3. Find the values of $f(a,b) - f(x,y)$ for points $(x,y)$ in a small circle around $(a,b)$. 
   * If these values are ALL positive then $(a,b)$ is a local maximum
   * If these values are ALL negative then $(a,b)$ is a local minimum
   * If some values are positive and some values are negative, then $(a,b)$ is a saddle point.

Here is some R code that you can use to check values of $f(x,y)$ in a small circle around a point $(a,b)$.

```{}
r = 0.1
theta = seq(0,2*pi,pi/10)
f(a,b) - f(a+r*cos(theta), b+r*sin(theta)) 
```


## Activities


### Characterize the Extrema

Multiple 2D functions have been (contour) plotted below. For each, identify the critical points and determine if they are maximums, minimums, or saddle points. 

```{r fig.align="center"}
f = makeFun(x^2+y^2+3~x&y)
plotFun(f(x,y)~x&y, xlim=range(-2,2), ylim=range(-2,2), filled=FALSE)
```

```{r fig.align="center"}
f = makeFun(-x*y*exp(-x^2-y^2)~x&y)
plotFun(f(x,y)~x&y, xlim=range(-1.5,1.5), ylim=range(-1.5,1.5), filled=FALSE)
```

```{r fig.align="center"}
f = makeFun(y^2-x^2~x&y)
plotFun(f(x,y)~x&y, xlim=range(-2,2), ylim=range(-2,2), filled=FALSE)
```

```{r fig.align="center"}
f = makeFun(cos(x)*(x^2-y^2)~x&y)
plotFun(f(x,y)~x&y, xlim=range(-1.5,1.5), ylim=range(-1.5,1.5), filled=FALSE)
```

## Classifying a critical point using a small circle of values

The function $f(x,y) =3x^2y + y^3 -3x^2-3y^2+3$ has four critical points:
$$
(0,0) \qquad (0,2) \qquad (1,1) \qquad (1,,-1).
$$
For each critical point $(a,b)$, evaluate $f(a,b) - f(x,y)$ in a small circle centered at $(a,b)$ to determine whether it is a local minimum, a local maximum, or a saddle point


## The 2D Second Derivative Test

The function $f(x,y)=10 + x^2 + 8y^2 - 3xy$ has two critical points: $(0,0)$ and $(1/2,1/4)$. 

1. Use the second derivative test to classify each point as a local minimum, local maximum, or saddle point. For convenience, here are the second derivatives of $f(x,y)$.

$$
f_{xx} = 6x, \qquad f_{yy} = 48y, \qquad f_{xy} = -3.
$$

2. Confirm that your answer is correct by creating a contour plot on a very small neighborhood around each critical point.


## Classify the Critical Points

Find the critical points of $f(x,y)=9-2x+4y-x^2-4y^2$. Then use any method that you like to classify each one as a local minimum, a local maximuum or a saddle point.

## Optimizing Flight Control

The range over which a flying drone can be guided is related to the atmospheric temperature and humidity. The range function is 
$$
R(t,h)=27800-5t^2-6ht-3h^2+400t+300h
$$
What are the optimal atmospheric conditions?